Online learning methods for financial time series forecasting

### 2017-11-25 slides/defense-slides.tex
Drunk example ready
partial final draft

\documentclass{beamer}
\usetheme{Warsaw}
%\usetheme{metropolis}
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{adjustbox}
\usepackage{color}
\usepackage{tikz}
\usetikzlibrary{shapes,arrows}
\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

@@

A weakly stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its error term in a particular period of time. For stationary time series, shocks to the system will gradually die away. 
\end{frame}

@@

\begin{columns}
\begin{column}{.4\linewidth}<1->
\begin{center}
        \begin{tikzpicture}[overlay]
            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
        \end{tikzpicture}
\end{center}
\end{column}
\begin{column}{.08\linewidth}<2->
        \begin{tikzpicture}[overlay]
            \node[yshift=-50] (img2) {\includegraphics[scale=0.06]{img/dog}};
        \end{tikzpicture}
\end{column}
%\begin{column}{.5\linewidth}<3->
%    \begin{center}
%        \begin{tikzpicture}[overlay]
%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
%        \end{tikzpicture}
%    \end{center}
%\end{column}
\end{columns}
\end{frame}

\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
 \begin{itemize}
  \item The drunk and dog's walk
 \end{itemize}
    \begin{center}
        \begin{tikzpicture}[overlay]
            \node[yshift=-60] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
        \end{tikzpicture}
    \end{center}
\end{frame}

@@

@@

\begin{frame}
\begin{alertblock}{VECM limitation}
The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data.
\end{alertblock}
\end{frame}

\begin{frame}
\frametitle{Online learning}
\begin{figure}
\includegraphics[width=0.8\paperwidth]{img/onlinelearning}
\end{figure}
\end{frame}

\subsection{Proposal}
\begin{frame}
\frametitle{Proposal}
\begin{block}{Proposal}
To use VECM with high frequency data where the best parameters are updated when new data is available and do this is done in a fast way.
\end{block}
\end{frame}

%\section{Background}
%
%
%\subsection{Cointegration}

%
%\subsection{Vector Autorregresive models}
%\begin{frame}
%\frametitle{VECM: Vector Error Correction model}
%
%VECM is a special form of a Vector Autorregresive model (VAR) for I(1)
%variables $\mathbf{y}$ that are also cointegrated. Let 
%
%\begin{equation*}
%\label{eq:variables}
%\mathbf{y}_t = 
%\begin{bmatrix} y_{1,t} & y_{2,t} & \dots & y_{l,t}
%\end{bmatrix}^\top \, \qquad t=p+1, \dots N. 
%\end{equation*}
%
%\begin{block}{VAR}
%VAR($p$) model describes the behaviour of a set of $l$
%endogenous variables as a linear combination of their last $p$ values:
%{\color{blue}
%\begin{equation*}
%\label{eq:var}
% \mathbf{y}_t = \phi_1 \mathbf{y}_{t-1}  + \dots +   \phi_p\mathbf{y}_{t-p}
%+ \mathbf{c} + \mathbf{\epsilon}_t \, ,
%\end{equation*}}
%\noindent where  ${\phi_1,\dots,\phi_p}$ are $l \times l$
%matrices of real coefficients,
%$\mathbf{\epsilon}_{p+1},\dots,\mathbf{\epsilon}_N$ are error terms,
%$\mathbf{c}$ is a constant vector and $N$ is the total number of samples.
%\end{block}
%\end{frame}
%
%
%

%
%\noindent where 
%%coefficients matrices $\Omega$ and $\phi_i^*$ are
%%function of matrices $\phi_i$ as follows:
%
%\begin{eqnarray*}
%\phi_i^* &: =& -\sum_{j=i+1}^{p} \phi_j \\
%\Omega &: =& -(\mathbb{I}-\phi_1-\dots-\phi_p) 
%\end{eqnarray*}
%\end{frame}
%
%
%
%\begin{frame}
%\frametitle{VECM matrix form}
%\small
%\begin{equation*}
%\underbrace{
%      \begin{bmatrix}
%       \mathbf{\Delta y}_{p+1}  \\ 
%       \mathbf{\Delta y}_{p+2}  \\ 
%       \vdots                   \\ 
%       \mathbf{\Delta y}_N      
%      \end{bmatrix}}_{\mathbf{B} } =
%\arraycolsep=1.4pt  
%\underbrace{\begin{bmatrix}
%   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
%  \end{bmatrix}}_{\mathbf{X}^\top}
%\underbrace{\begin{bmatrix}
% \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
% \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
% \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
% \vdots                  & \vdots                 & \ddots&  \vdots                   \\
% \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
% 1                      & 1                       & \dots     & 1   
% \end{bmatrix}}_{\mathbf{A}^\top}
%+
%\underbrace{\begin{bmatrix}
%              \mathbf{\epsilon}_{p+1} \\ 
%              \mathbf{\epsilon}_{p+2} \\ 
%              \vdots \\ 
%              \mathbf{\epsilon}_N
%             \end{bmatrix}}_{\mathbf{E}^\top}
%\end{equation*}
%\end{frame}
%
%\section{Proposal}
%\subsection{Motivation}
%\begin{frame}
%\frametitle{VECM difficulties in an online context}
%\begin{itemize}
%\item VECM assumes cointegration vectors do not change in time. However, these might change due to several economic factors.
%\item To get updated cointegration vectors in time is computationally expensive because 
%the Johansen method and OLS are used to obtain them. It would be good to know how the number of cointegration vectors are affected by the number of data used $L$ and the number of lags of VECM $p$.
%\end{itemize}
%\end{frame}
%
%
%\begin{frame}
%\frametitle{Data}
%\begin{itemize}
%\item Tests were carried out using four foreign exchange rates all related to
%USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
%\item This data was collected from Dukascopy, a free
%database which gives access to the Swiss Foreign Exchange marketplace.
%\item The tests were done using 10-seconds frequency from ask prices
%from the 11th to the 15th of August 2014. 
%%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
%\end{itemize}
%\end{frame}
%
%
%\begin{frame}
%\frametitle{Cointegration vectors distribution}
%\begin{figure}[!h]
%  %\vspace{-0.8cm}
%  \centering
%  \includegraphics[width=0.75\textwidth]{img/histCointVectors-offset20520-p-1-freq-10s}
%  \caption{Histogram of the number of cointegration vectors using $p=1$. Four
%  possible values for $L$ are shown (100, 2000, 3000 and 4000).}
%  \label{fig:hists}
%\end{figure}
%\end{frame}
%
%\begin{frame}
%\frametitle{Percentage of cointegration}
%Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
%The interesting cases of cointegration are those where $r$ lies strictly
%between $0$ and $l$, i.e. $0<r<l$.
%\begin{block}{Percentage of cointegration (PC)}
%{\color{blue}
%\begin{equation*} \label{eq:pcoint}
%PC = 
%\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
%     {\#it}\times 100
%\end{equation*}}
%\end{block}
%\end{frame}
%
%
%\begin{frame}
%\frametitle{PC vs MSE}
%\begin{figure}[ht!]
%  %\vspace{-0.8cm}
%  \centering
%  \includegraphics[width=0.7\textwidth]{img/MSE-offset20520-p-2-freq-10s}
%  \caption{MSE versus the percentage of cointegration considering 1000
%  iterations. }
%  \label{fig:cointvsmse}
%\end{figure}
%\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
%\subsection{AVECM}
%\begin{frame}
%\frametitle{Proposal: Adaptive VECM algorithm (AVECM)}
%\begin{itemize}
%\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
%cointegration $PC$ in the near past. This process was done every time that new data was processed. 
%\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
%\end{itemize}
%\end{frame}
%
%
%%\begin{frame}
%%\frametitle{AVECM Algorithm}
%%\begin{algorithmic}[1]
%%\REQUIRE $\,$ \\
%%$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
%%$j$: Starting point of testing \\
%%$it$: Ending point of testing \\
%%$ps$: list of $p$ values \\
%%$Ls$: list of $L$ values ($L<N$) \\
%%$m$: Iterations to determine parameters ($m < N-L$)\\
%%\ENSURE  $\,$ \\
%%$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
%%\FOR { $i =j$ to $it$ }
%%   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
%%    \STATE $L,p \gets
%%    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
%%        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
%%        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
%%\ENDFOR
%%\end{algorithmic}
%%\end{frame}
%
%\subsection{Model comparison} 
%\begin{frame}
%\frametitle{Random walk}
%The random walk model is defined as:
%\begin{equation}
%\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
%\label{rwmodel}
%\end{equation}
%The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
%\begin{equation}
%\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
%\end{equation}
%\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
%\end{frame}
%
%
%\begin{frame}
%\frametitle{ARIMA}
%A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
%\begin{equation}
%\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
%\end{equation}
%\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
%\end{frame}
%
%\begin{frame}
%\frametitle{Results}
%\begin{tabular}{ccccccc}
%
%& \multicolumn{3}{l}{MSE} & &
%\multicolumn{2}{c}{$U$-statistic} \\ 
%\hhline{~---~--} \\
% &
%AVECM & ARIMA & p-value & &
%AVECM & ARIMA \\ 
%
%\hline
% EURUSD & 1.07 e-09 & 1.14 e-09 &  9.25 e-12 & & 0.6863 & 0.7108\\
% GBPUSD & 1.66 e-09 & 1.74 e-09 &  6.95 e-02 & & 0.6866 & 0.7025\\
% USDCHF & 5.85 e-10 & 6.35 e-10 &  2.89 e-14 & & 0.6803 & 0.7091\\
% USDJPY  & 6.34 e-06 & 6.51 e-06 &  6.85 e-05 & & 0.6964 & 0.7057\\
% \end{tabular}
%\end{frame}
%
%%\begin{frame}
%%\frametitle{Evaluation methods}
%%\begin{block}{MSE}
%%Measures the distance between forecasts
%%and the true values and large deviations from the true value have a
%%large impact due to squaring forecast error.
%%{\color{blue}
%%\begin{equation}\label{eq:MSE}
%%\text{MSE} = 
%%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
%%\end{equation}}
%%\end{block}
%%\begin{block}{\bf $U$-statistic}
%%Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
%%\end{block}
%%\end{frame}
%
%\section{Experiments}
%\subsection{Uno}
%
%%\begin{frame}
%%\frametitle{Unit root tests}
%%VECM can only be used if the time series are I(1). Therefore, 
%%before running the tests, we firstly checked whether the time series were
%%I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
%%\end{frame}
%%
%%\begin{frame}
%%\frametitle{Performance accuracy}
%%\small
%%\begin{tabular}{ccccccc}
%%& \multicolumn{3}{l}{MSE} & &
%%\multicolumn{2}{c}{$U$-statistic} \\ 
%%\hhline{~---~--} \\
%% & AVECM & ARIMA & p-value & &
%%AVECM & ARIMA \\ 
%%
%%\hline
%% EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
%% GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
%% USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
%% USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
%% \end{tabular}
%% \end{frame}
%%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.
%
%
%
%\begin{frame}
%\frametitle{Parallel implementation}
%The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
%with wide portability to be built, being able to run in large clusters
%or on local computers.
%We tested our proposal in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
%of 24 cores each (48 cores in total) and 24GB RAM.
%\end{frame}
%
%\begin{frame}
%\frametitle{Parameters setting}
%\begin{itemize}
%%\item Parameter $it=100$
%\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
%between 1 and 5. 
%\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
%\end{itemize}
%\end{frame}
%
%\begin{frame}
%\frametitle{Execution times}
%\begin{figure}[ht]
%  \centering

>>>>

\documentclass{beamer}
\usetheme{Warsaw}
%\usetheme{metropolis}
%\usepackage[spanish,activeacute]{babel}
\usepackage[utf8]{inputenc}
\usepackage{listings}
\usepackage{amsmath,amsfonts,amssymb}
\usepackage{adjustbox}
\usepackage{color}
\usepackage{tikz}
\usepackage{array}
\usepackage{booktabs}
\usetikzlibrary{shapes,arrows}
\usepackage{multicol}
\usepackage{hhline}
\usepackage{pgf}
\usepackage{algorithm} 
\usepackage{algorithmic}
\renewcommand{\algorithmicrequire}{\textbf{Input:}}
\renewcommand{\algorithmicensure}{\textbf{Output:}}
\def\R{{\mathbb R }}
\def\N{{\mathbb N }}

@@

A weak stationary time series is a process which mean, variance and auto covariance do not change over time:
\small
{\color{blue}
 \begin{eqnarray*} E(y_t) &=& \mu  \quad
\forall t \in \mathbb{N} \\ E(y^2_t) &=& \sigma^2  \quad \forall t \in
\mathbb{N} \\ \lambda(s,t)&=&\lambda(s+h,t+h) \quad \forall s,t \in \mathbb{N},
\forall h \in \mathbb{Z} \end{eqnarray*}}
\end{block}
One of the consequences of stationary processes is how they recover from a shock. A shock represents an unexpected change in a variable or a in its error term in a particular period of time. For stationary time series, shocks to the system will gradually die away. 
\end{frame}

@@

\begin{columns}
\begin{column}{.4\linewidth}<1->
\begin{center}
        \begin{tikzpicture}[overlay]
            \node[yshift=-50] (img2) {\includegraphics[scale=1.4]{img/drunk}};
        \end{tikzpicture}
\end{center}
\end{column}
\begin{column}{.08\linewidth}<2->
        \begin{tikzpicture}[overlay]
            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/dog}};
        \end{tikzpicture}
\end{column}
%\begin{column}{.5\linewidth}<3->
%    \begin{center}
%        \begin{tikzpicture}[overlay]
%            \node[yshift=-50] (img2) {\includegraphics[scale=0.1]{img/mananddog}};
%        \end{tikzpicture}
%    \end{center}
%\end{column}
\end{columns}
\end{frame}

\begin{frame}[t,fragile]{Random walk and Cointegration Illustration}
 \begin{itemize}
  \item The drunk and dog's walk
 \end{itemize}
    \begin{center}
        \begin{tikzpicture}[overlay]
            \node[yshift=-60] (img2) {\includegraphics[scale=0.7]{img/mananddog}};
        \end{tikzpicture}
    \end{center}
\end{frame}

@@

\begin{frame}
\frametitle{VECM matrix form}
\small
\begin{equation*}
\underbrace{
      \begin{bmatrix}
       \mathbf{\Delta y}_{p+1}  \\ 
       \mathbf{\Delta y}_{p+2}  \\ 
       \vdots                   \\ 
       \mathbf{\Delta y}_N      
      \end{bmatrix}}_{\mathbf{B} } =
\arraycolsep=1.4pt  
\underbrace{\begin{bmatrix}
   \alpha^\top & \phi^*_1 & \dots & \phi^*_{p-1} &\mathbf{c}   
  \end{bmatrix}}_{\mathbf{X}^\top}
\underbrace{\begin{bmatrix}
 \beta\mathbf{y}_p      & \beta\mathbf{y}_{p+1}   & \dots&   \beta\mathbf{y}_{N-1}   \\
 \mathbf{\Delta y}_{p}   & \mathbf{\Delta y}_{p+1}& \dots &  \mathbf{\Delta y}_{N-1} \\
 \mathbf{\Delta y}_{p-1} & \mathbf{\Delta y}_{p}  & \dots &  \mathbf{\Delta y}_{N-2}   \\
 \vdots                  & \vdots                 & \ddots&  \vdots                   \\
 \mathbf{\Delta y}_{2}   & \mathbf{\Delta y}_{3} & \dots &   \mathbf{\Delta y}_{N-p+1} \\
 1                      & 1                       & \dots     & 1   
 \end{bmatrix}}_{\mathbf{A}^\top}
+
\underbrace{\begin{bmatrix}
              \mathbf{\epsilon}_{p+1} \\ 
              \mathbf{\epsilon}_{p+2} \\ 
              \vdots \\ 
              \mathbf{\epsilon}_N
             \end{bmatrix}}_{\mathbf{E}^\top}
\end{equation*}
\end{frame}

@@

\begin{frame}
\frametitle{VECM limitations}

The Vector Error correction model (VECM) is used to model cointegrated time series but only with batch data and rarely used with high frequency data  mainly due to computational limitations:

\begin{itemize}
\item Calculation of the cointegration vectors
\item VECM parameters are obtained using the ordinary least squares (OLS) method.
\end{itemize}

\section{Proposal}
\begin{frame}
\frametitle{Proposal}
\begin{block}{Proposal}
To adapt VECM to be used with high frequency data, where the best parameters are updated when new data.
\end{block}

\begin{block}{Hypothesis}
An algorithm based on cointegration and high
performance computing will allow faster forecasting
algorithms for financial time series to be obtained while maintaining good accuracy levels.
\end{block}
\end{frame}

\section{First Contribution}
\begin{frame}
\frametitle{First approach}
To use VECM with high frequency data, where optimised parameters are obtained when new data. The optimisation was made in a distributed environment.
\end{frame}

\subsection{Data}
\begin{frame}
\frametitle{Data}
\begin{itemize}
\item Tests were carried out using four foreign exchange rates all related to
USD: EURUSD, GBPUSD, USDCHF and USDJPY. 
\item This data was collected from Dukascopy, a free
database which gives access to the Swiss Foreign Exchange marketplace.
\item The tests were done using 10-seconds frequency from ask prices
from the 11th to the 15th of August 2014. 
%\item Since one day corresponds to 8640 data points and we used 5 days of data, we have 43,200 data points in total.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Cointegration vectors distribution}
  \begin{figure}[!h]
  %\vspace{-0.8cm}
  \centering
   \includegraphics[width=0.65\textwidth]{img/51_Fig1}
  \caption[Distribution of the number of cointegration vectors using $p=1$ lags]{Distribution of the number of cointegration vectors using $p=1$ lags.
  Four possible values for windows size $L$ are shown (2, 6, 10 and 14) hours (1
  hour = 360 data points).}
  \label{fig:hists}
\end{figure}
\end{frame}

\begin{frame}
\frametitle{Percentage of cointegration}
Since $r=0$ means no cointegration and $r=l=4$ reveals that no process is I(1) but stationary.
The interesting cases of cointegration are those where $r$ lies strictly
between $0$ and $l$, i.e. $0<r<l$.
\begin{block}{Percentage of cointegration (PC)}
{\color{blue}
\begin{equation*} \label{eq:pcoint}
PC = 
\frac{\#\{ it \mid \text{$it$ has $r$ c.v. with $0<r<l$}\}}
     {\#it}\times 100
\end{equation*}}
\end{block}
\end{frame}

%\begin{frame}
%\frametitle{PC vs MSE}
%\begin{figure}[ht!]
%  %\vspace{-0.8cm}
%  \centering
%  \includegraphics[width=0.7\textwidth]{img/MSE-offset20520-p-2-freq-10s}
%  \caption{MSE versus the percentage of cointegration considering 1000
%  iterations. }
%  \label{fig:cointvsmse}
%\end{figure}
%\end{frame}
%%%%%%%%%%%%%%%%%%%%%%%%%%%
%
\subsection{AVECM}
\begin{frame}
\frametitle{Adaptive VECM algorithm (AVECM)}
\begin{itemize}
\item We proposed to choose $L$ and $p$ in order to maximise the percentage of
cointegration $PC$ in the near past. This process was done every time that new data was processed. 
\item To use a distributed environment to make this parameters search, since this is the most expensive routine.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{AVECM Algorithm}
\small
\begin{algorithmic}[1]
\REQUIRE $\,$ \\
$\mathbf{y}$: matrix with $N$ input vectors and $l$ time series\\
$j$: Starting point of testing \\
$it$: Ending point of testing \\
$ps$: list of $p$ values \\
$Ls$: list of $L$ values ($L<N$) \\
$m$: Iterations to determine parameters ($m < N-L$)\\
\ENSURE  $\,$ \\
$\{ \hat{\mathbf{y}}[1],\dots,\hat{\mathbf{y}}[it]\}$: prediction vectors \\
\FOR { $i =j$ to $it$ }
   \STATE $\mathbf{Y} \gets \mathbf{y}[:,i-1]$
    \STATE $L,p \gets
    \texttt{get\_best\_params}(Ls,ps,m,\mathbf{Y})$
        \STATE $model = \text{VECM}(\mathbf{Y},L, p)$
        \STATE $\hat{\mathbf{y}}[i-j] = model.predict()$
\ENDFOR
\end{algorithmic}
\end{frame}

%\subsection{Model comparison} 
%\begin{frame}
%\frametitle{Random walk}
%The random walk model is defined as:
%\begin{equation}
%\mathbf{y}_t = \mathbf{y}_{t-1} + \epsilon_{t}
%\label{rwmodel}
%\end{equation}
%The naive forecast of the time series difference $\hat{\mathbf{y}}_{t+1}$ for the random walk model is defined as:
%\begin{equation}
%\hat{\mathbf{y}}_{t+1} = \mathbf{y}_t + \hat{\epsilon}_{t+1} 
%\end{equation}
%\noindent where  $\hat{\epsilon}_{t+1} = \epsilon_{t}$.
%\end{frame}
%
%
%\begin{frame}
%\frametitle{ARIMA}
%A process can be modelled as an ARIMA$(p,d,q)$ if $\mathbf{x}_t = \Delta^d \mathbf{y}_t $, is an ARMA$(p,q)$. An ARMA$(p,q)$ model is the following:
%\begin{equation}
%\mathbf{x}_t = \sum_{i=1}^p \phi_i \mathbf{x}_{t-i}  +  \sum_{j=1}^q \theta_j \epsilon_{t-j}  
%\end{equation}
%\noindent with coefficients $\phi_p \neq 0$, $\theta_q \neq 0$ and $\sigma_{\epsilon}^2 > 0$.
%\end{frame}

%%\begin{frame}
%%\frametitle{Evaluation methods}
%%\begin{block}{MSE}
%%Measures the distance between forecasts
%%and the true values and large deviations from the true value have a
%%large impact due to squaring forecast error.
%%{\color{blue}
%%\begin{equation}\label{eq:MSE}
%%\text{MSE} = 
%%\frac{\displaystyle \sum_{t=1}^{N} (\mathbf{y}_t-\hat{\mathbf{y}}_t)^2}{N}
%%\end{equation}}
%%\end{block}
%%\begin{block}{\bf $U$-statistic}
%%Unit free measure obtained as the ratio between the root MSE (RMSE) of a model and the RMSE of the naive random walk model. 
%%\end{block}
%%\end{frame}
%
%\section{Experiments}
%\subsection{Uno}
%
\begin{frame}
\frametitle{Unit root tests}
VECM can only be used if the time series are I(1). Therefore, 
before running the tests, we firstly checked whether the time series were
I(1) using the Augmented Dickey Fuller (ADF) test with lags $p=1,2,3,4,5$.
\begin{table}[ht]
\centering
\tiny
\begin{tabular}{llllll}
\toprule
{Variable} & {ADF(1)} & {ADF(2)} & {ADF(3)} & {ADF(4)} & {ADF(5)}\\ 
\midrule
EURUSD &  -0.052   & -0.054  & -0.054  & -0.054  & -0.054  \\
GBPUSD &  -0.744  & -0.784  & -0.805  & -0.837  & -0.846  \\
USDCHF &  -0.476   & -0.493  & -0.493  & -0.495  & -0.502  \\
USDJPY &  0.357   & 0.360  & 0.360  & 0.367  & 0.367  \\
$\Delta$ EURUSD & -128.4*  & -128.4*  & -96.85* & -89.12*   & -89.12*\\
$\Delta$ GBPUSD & -131.4*  & -112.7*  & -102.5* & -92.86*   & -88.29*\\
$\Delta$ USDCHF & -127.8*  & -127.8*  & -96.94* & -88.82*   & -80.79*\\
$\Delta$ USDJPY & -135.1*  & -135.1*  & -101.2* & -101.2*   & -101.2*\\
\bottomrule
\addlinespace[1ex]
\multicolumn{6}{l}{ \textsuperscript{*} Indicates significance at 1\% level}
\end{tabular}
\caption{Unit roots tests for EURUSD, GBPUSD, USDCHF and USDJPY at 10-second
frequency.}
\label{tab:adf}
\end{table}
\end{frame}

\begin{frame}
\frametitle{Performance accuracy}
\small
\begin{tabular}{ccccccc}
& \multicolumn{3}{l}{MSE} & &
\multicolumn{2}{c}{$U$-statistic} \\ 
\hhline{~---~--} \\
 & AVECM & ARIMA & p-value & &
AVECM & ARIMA \\

\hline
 EURUSD & 1.0702 e-09 & 1.1481 e-09 &  9.250 e-12 & & 0.6863 & 0.7108\\
 GBPUSD & 1.6630 e-09 & 1.7408 e-09 &  6.951 e-02 & & 0.6866 & 0.7025\\
 USDCHF & 5.8503 e-10 & 6.3545 e-10 &  2.899 e-14 & & 0.6803 & 0.7091\\
 USDJPY  & 6.3483 e-06 & 6.5194 e-06 &  6.853 e-05 & & 0.6964 & 0.7057
 \end{tabular}
 \end{frame}
%%For AVECM we considered different number of iterations (parameter $m$ in algorithm \ref{alg:AVECM}): 10, 50 and 100. We tried 12, 24 and 47 different pair of combinations for $L$ and $p$. Possible values for $L$ were in the interval $[100,4000]$ and $p$ can have values in between $[1,5]$. Best AVECM performance was compared against ARIMA and the random walk model.
%%Table~\ref{tab:stats} shows the out-of-sample performance measures: MSE and $U$-statistic for AVECM and ARIMA. In terms of both measures we found that AVECM is superior to ARIMA and the naive random walk model. We also included the p-value that proves that the difference in the MSE is significant at the 99\% significance level in three of the four currency rates and at 90\% in the case of GBPUSD. The $U$-statistic shows that AVECM and ARIMA are superior to the naive random walk model and that our proposal is also superior to ARIMA.
%
%
%
\begin{frame}
\frametitle{Parallel implementation}
The function that get best parameters $L$ and $p$ was implemented using MPI in Python. MPI allows large-scale parallel applications
with wide portability to be built, being able to run in large clusters
or on local computers.
Tests ran in a cluster with 2 servers Xeon E5-2667 (2.90GHz)
of 24 cores each (48 cores in total) and 24GB RAM.
Parameter settings:
\begin{itemize}
%\item Parameter $it=100$
\item The $L$ parameter was always chosen between 100 and 4000 and $p$ always took values
between 1 and 5. 
\item Parameter $nparams$ represents the number of pairs ($L$,$p$) used to maximise the percentage of cointegration. 
\end{itemize}
\end{frame}
%
\begin{frame}
\frametitle{Execution times}
\begin{figure}[ht]
  \centering
  \includegraphics[width=0.7\textwidth]{img/51_Fig3}
  \caption[Computing time and Speed-up]{Computing time of sequential and parallel algorithm is shown in the
  upper figure. Speed-up is shown below.}
  \label{fig:extimes}
\end{figure}
\end{frame}

\section{Conclusions}
\subsection{Uno}
\begin{frame}
\frametitle{Conclusions}
\begin{itemize}
\item Cointegration relations change with time. We empirically showed that the Johansen method is sensitive to the number of lags but also to the amount of data considered.
\item We found that out-of-sample forecast performance MSE is related to the {\em percentage of cointegration\/}.  
\item VECM improves performance measures by finding
parameters of $L$ and $p$ maximising the percentage of cointegration.
\item Increasing the number of parameters will always lead to
better performance measures.
\item The parallel implementation allowed the execution times to be reduced
more than 9 times and therefore a response time was obtain before 10
seconds. 
\end{itemize}
\end{frame}

\section{Second Contribution}

\begin{frame}
\frametitle{Second approach}
 Recently, online learning algorithms have been proposed to solve problems with large data sets because of their simplicity and their ability to update the model when new data is
available.
\begin{figure}
\includegraphics[width=0.6\paperwidth]{img/onlinelearning}
\end{figure} 
\end{frame}

\begin{frame}
\frametitle{Proposal: Online VECM algorithm (OVECM)}
\begin{itemize}
\item Considers only a sliding window of the $L$ most recent data. 
\item Since cointegration vectors represent long-run relationships they vary
little in time. OVECM obtains these vectors only when they need to be updated. 
\item This updating is done when MAPEs of the last $n$ inputs is above a certain
error given by a threshold 
\item OVECM also implements matrix optimisations in order to reduce execution time,
such as updating matrices with new data, removing old data and introducing new
cointegration vectors.
\end{itemize}
\end{frame}

\begin{frame}
\frametitle{Methodology}
OVECM were compared against VECM and ARIMA. VECM and ARIMA are called with a
sliding window of the most recent data, whereby the models are updated at every
time step. Therefore, we named them SLVECM and SLARIMA.
\end{frame}

\subsection{Parameters}
\begin{frame}
\frametitle{Parameters optimization}
VECM order and ARIMA parameters were selected using AIC.

\begin{table}[ht]
\label{tab:params}
\begin{center}
\begin{tabular}{|c|c|c|}
\hline
Windows size $L$ & VECM & ARIMA\\
 $L$ & order $(p)$ & order $(p,d,q)$ \\
\hline
100 & 2 & p=2,d=1,q=1\\
400 & 5 & p=1,d=1,q=1\\
700 & 3 &p=2,d=1,q=1\\
1000 &3 & p=2,d=1,q=1\\
\hline
\end{tabular}
\end{center}
\end{table}
\end{frame}
